---
title: "CDC Data Exercise"
author: Austin Thrash
date: today
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    highlight-style: github
---

## Importing packages

```{r}
library(here)
library(readr)
library(gridExtra)
library(grid)
library(ggplot2)
library(dplyr)
```

## Importing data

```{r}
# This chunk is used to import that data.

# Define the location of the data file
data_location <- here::here("cdcdata-exercise", "cdcdata_sample.csv")

# Read the CSV file into a dataframe
cdcdata <- read_csv(data_location)

# View the first few rows of the dataframe
head(cdcdata)
```

# Exploring the data

The dataset I selected is '**Provisional COVID-19 Deaths by Sex and Age**' from the CDC website. The dataset details deaths involving COVID-19, pneumonia, and influenza reported to NCHS by sex, age group, and jurisdiction of occurrence. I took a sample of this dataset, it originally had 138K entries that were grouped by how they were collected (by month, by year, by total), I took the 'By Total' group and dropped NA entries, in turn creating a dataset with 1305 entries.

Below is a table describing the variables I will be using from the dataset:

| Variable                                 | Description                                   |
|------------------------------------------|-----------------------------------------------|
| State                                    | Jurisdiction of occurence                     |
| Sex                                      | Sex                                           |
| Age Group                                | Age Group                                     |
| COVID-19 Deaths                          | Deaths involving COVID-19                     |
| Pneumonia Deaths                         | Pneumonia Deaths                              |
| Influenza Deaths                         | Influenza Deaths                              |
| Pneumonia, Influenza, or COVID-19 Deaths | Deaths with Pneumonia, Influenza, or COVID-19 |
| Total Deaths                             | Deaths from all causes of death               |

Link:

[CDC Dataset Source](https://data.cdc.gov/NCHS/Provisional-COVID-19-Deaths-by-Sex-and-Age/9bhg-hcku/about_data)

## Cleaning data

```{r}
# This chunk will be used to clean the data. The data was cleaned prior, we really just need to drop the columns we arent using.

# Removing unused columns
cdcdata$`Data As Of` <- NULL
cdcdata$`Start Date` <- NULL
cdcdata$`End Date` <- NULL
cdcdata$Group <- NULL

names(cdcdata) <- make.names(names(cdcdata))
```

## Exploratory Analysis

### Summary()

```{r}
summary(cdcdata)
```

### Distributions

plots should include percentages if categorical, for continuous, plot and then summarize its mean and standard dev

```{r}
# Calculate counts and percentages for Age Group
cdcdata_summary <- cdcdata %>%
  group_by(Age.Group) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

# Plot with percentage labels
ggplot(cdcdata_summary, aes(x = Age.Group, y = count)) + 
  geom_bar(stat = "identity", fill = "salmon", color = "black") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), vjust = -0.5) +
  labs(title = "Age Group", x = "Age Group", y = "Count") +
  theme_minimal()
```

```{r}
# Calculate counts and percentages for Sex
cdcdata_summary <- cdcdata %>%
  group_by(Sex) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

# Plot with percentage labels
ggplot(cdcdata_summary, aes(x = Sex, y = count)) + 
  geom_bar(stat = "identity", fill = "salmon", color = "black") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), vjust = -0.5) +
  labs(title = "Sex", x = "Sex", y = "Count") +
  theme_minimal()
```

```{r}
ggplot(cdcdata, aes_string(x = "Pneumonia..Influenza..or.COVID.19.Deaths")) + 
        geom_bar(fill = "salmon", color = "black") +
        labs(title = "Pneumonia, Influenza, or COVID-19 Deaths") +
        theme_minimal()
```

```{r}
ggplot(cdcdata, aes_string(x = "Influenza.Deaths")) + 
        geom_histogram(binwidth = 1, fill = "salmon", color = "black") +
        labs(title = "Influenza Deaths") +
        theme_minimal()
```

```{r}
ggplot(cdcdata, aes_string(x = "Pneumonia.Deaths")) + 
        geom_bar(fill = "salmon", color = "black") +
        labs(title = "Pneumonia Deaths") +
        theme_minimal()
```

```{r}
ggplot(cdcdata, aes_string(x = "COVID.19.Deaths")) + 
        geom_bar(fill = "salmon", color = "black") +
        labs(title = "COVID-19 Deaths") +
        theme_minimal()
```

```{r}
ggplot(cdcdata, aes_string(x = "Total.Deaths")) + 
        geom_bar(fill = "salmon", color = "black") +
        labs(title = "Total Deaths") +
        theme_minimal()
```

```{r}

mean(cdcdata$Pneumonia..Influenza..or.COVID.19.Deaths)
sd(cdcdata$Pneumonia..Influenza..or.COVID.19.Deaths)
```

```{r}
mean(cdcdata$Influenza.Deaths)
sd(cdcdata$Influenza.Deaths)
```

```{r}
mean(cdcdata$Pneumonia.Deaths)
sd(cdcdata$Pneumonia.Deaths)
```

```{r}
mean(cdcdata$COVID.19.Deaths)
sd(cdcdata$COVID.19.Deaths)
```

```{r}
mean(cdcdata$Total.Deaths)
sd(cdcdata$Total.Deaths)
```

## Summary of data for Part 2

| Variable                                 | Mean     | Standard Deviation |
|------------------------------------------|----------|--------------------|
| Pneumonia..Influenza..or.COVID.19.Deaths | 11629.85 | 66245.64           |
| Influenza.Deaths                         | 150.4058 | 826.2635           |
| Pneumonia.Deaths                         | 7705.152 | 43860.05           |
| COVID.19.Deaths                          | 7596.808 | 43164.85           |
| Total.Deaths                             | 81653.12 | 459491.2           |

I had some issues finding a dataset that I wanted to use, if this dataset gives you any trouble when attempting to replicate it synthetically, let me know.

# Making synthetic data

```{r}

```
