---
title: "CDC Data Exercise"
author: Austin Thrash
date: today
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    highlight-style: github
---

## Importing packages

```{r}
library(here)
library(readr)
library(gridExtra)
library(grid)
library(ggplot2)
library(dplyr)
```

## Importing data

```{r}
# This chunk is used to import that data.

# Define the location of the data file
data_location <- here::here("cdcdata-exercise", "cdcdata_sample.csv")

# Read the CSV file into a dataframe
cdcdata <- read_csv(data_location)

# View the first few rows of the dataframe
head(cdcdata)
```

# Exploring the data

The dataset I selected is '**Provisional COVID-19 Deaths by Sex and Age**' from the CDC website. The dataset details deaths involving COVID-19, pneumonia, and influenza reported to NCHS by sex, age group, and jurisdiction of occurrence. I took a sample of this dataset, it originally had 138K entries that were grouped by how they were collected (by month, by year, by total), I took the 'By Total' group and dropped NA entries, in turn creating a dataset with 1305 entries.

Below is a table describing the variables I will be using from the dataset:

| Variable                                 | Description                                   |
|------------------------------------------|-----------------------------------------------|
| State                                    | Jurisdiction of occurence                     |
| Sex                                      | Sex                                           |
| Age Group                                | Age Group                                     |
| COVID-19 Deaths                          | Deaths involving COVID-19                     |
| Pneumonia Deaths                         | Pneumonia Deaths                              |
| Influenza Deaths                         | Influenza Deaths                              |
| Pneumonia, Influenza, or COVID-19 Deaths | Deaths with Pneumonia, Influenza, or COVID-19 |
| Total Deaths                             | Deaths from all causes of death               |

Link:

[CDC Dataset Source](https://data.cdc.gov/NCHS/Provisional-COVID-19-Deaths-by-Sex-and-Age/9bhg-hcku/about_data)

## Cleaning data

```{r}
# This chunk will be used to clean the data. The data was cleaned prior, we really just need to drop the columns we arent using.

# Removing unused columns
cdcdata$`Data As Of` <- NULL
cdcdata$`Start Date` <- NULL
cdcdata$`End Date` <- NULL
cdcdata$Group <- NULL

names(cdcdata) <- make.names(names(cdcdata))
```

## Exploratory Analysis

### Summary()

```{r}
summary(cdcdata)
str(cdcdata)
```

### Distributions

plots should include percentages if categorical, for continuous, plot and then summarize its mean and standard dev

```{r}
# Calculate counts and percentages for Age Group
cdcdata_summary <- cdcdata %>%
  group_by(Age.Group) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

# Plot with percentage labels
ggplot(cdcdata_summary, aes(x = Age.Group, y = count)) + 
  geom_bar(stat = "identity", fill = "salmon", color = "black") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), vjust = -0.5) +
  labs(title = "Age Group", x = "Age Group", y = "Count") +
  theme_minimal()
```

```{r}
# Calculate counts and percentages for Sex
cdcdata_summary <- cdcdata %>%
  group_by(Sex) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

# Plot with percentage labels
ggplot(cdcdata_summary, aes(x = Sex, y = count)) + 
  geom_bar(stat = "identity", fill = "salmon", color = "black") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), vjust = -0.5) +
  labs(title = "Sex", x = "Sex", y = "Count") +
  theme_minimal()
```

```{r}
ggplot(cdcdata, aes_string(x = "Pneumonia..Influenza..or.COVID.19.Deaths")) + 
        geom_bar(fill = "salmon", color = "black") +
        labs(title = "Pneumonia, Influenza, or COVID-19 Deaths") +
        theme_minimal()
```

```{r}
ggplot(cdcdata, aes_string(x = "Influenza.Deaths")) + 
        geom_histogram(binwidth = 1, fill = "salmon", color = "black") +
        labs(title = "Influenza Deaths") +
        theme_minimal()
```

```{r}
ggplot(cdcdata, aes_string(x = "Pneumonia.Deaths")) + 
        geom_bar(fill = "salmon", color = "black") +
        labs(title = "Pneumonia Deaths") +
        theme_minimal()
```

```{r}
ggplot(cdcdata, aes_string(x = "COVID.19.Deaths")) + 
        geom_bar(fill = "salmon", color = "black") +
        labs(title = "COVID-19 Deaths") +
        theme_minimal()
```

```{r}
ggplot(cdcdata, aes_string(x = "Total.Deaths")) + 
        geom_bar(fill = "salmon", color = "black") +
        labs(title = "Total Deaths") +
        theme_minimal()
```

```{r}

mean(cdcdata$Pneumonia..Influenza..or.COVID.19.Deaths)
sd(cdcdata$Pneumonia..Influenza..or.COVID.19.Deaths)
```

```{r}
mean(cdcdata$Influenza.Deaths)
sd(cdcdata$Influenza.Deaths)
```

```{r}
mean(cdcdata$Pneumonia.Deaths)
sd(cdcdata$Pneumonia.Deaths)
```

```{r}
mean(cdcdata$COVID.19.Deaths)
sd(cdcdata$COVID.19.Deaths)
```

```{r}
mean(cdcdata$Total.Deaths)
sd(cdcdata$Total.Deaths)
```

## Summary of data for Part 2

| Variable                                 | Mean     | Standard Deviation |
|------------------------------------------|----------|--------------------|
| Pneumonia..Influenza..or.COVID.19.Deaths | 11629.85 | 66245.64           |
| Influenza.Deaths                         | 150.4058 | 826.2635           |
| Pneumonia.Deaths                         | 7705.152 | 43860.05           |
| COVID.19.Deaths                          | 7596.808 | 43164.85           |
| Total.Deaths                             | 81653.12 | 459491.2           |

I had some issues finding a dataset that I wanted to use, if this dataset gives you any trouble when attempting to replicate it synthetically, let me know.


# Making synthetic data
**This section contributed by Nicholas Stevenson**

To create the synthetic dataset the prompt I used was: "Create a dataset replicating an already exisiting dataset in R studio, The variables I'd like replicate is 'State'(character), 'Sex'(character), 'Age.Group'(character), 'Covid.19.Deaths'(number), 'Total.Deaths'(number), 'Pneumonia.Deaths'(number), 'Influenza.Deaths'(number). There are 1343 observations." 

I followed this was a series of descriptive statements by referencing the mean and standard deviation of the original data as well as referencing the original charts and whether they were normal or skewed. 
```{r}
set.seed(123) # for reproducibility

# Define the variables
states <- c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", 
            "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", 
            "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", 
            "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", 
            "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", 
            "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", 
            "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", 
            "Wisconsin", "Wyoming")
sexes <- c("Male", "Female", "All Sexes")
sex_probs <- c(0.315, 0.293, 0.392)
age_groups <- c("0-17", "18-29", "30-39", "40-49", "50-64", "65-74", "75-84", "85+")
age_group_probs <- c(0.05, 0.10, 0.10, 0.15, 0.20, 0.20, 0.15, 0.05) # Skewed probabilities
n <- 1343

# Generate the dataset
synthetic_data <- tibble(
  State = sample(states, n, replace = TRUE),
  Sex = sample(sexes, n, replace = TRUE, prob = sex_probs),
  Age.Group = sample(age_groups, n, replace = TRUE, prob = age_group_probs),
  Covid.19.Deaths = rnorm(n, mean = 7596.808, sd = 43164.85),
  Total.Deaths = rnorm(n, mean = 81653.12, sd = 45491.2),
  Pneumonia.Deaths = rnorm(n, mean = 7705.152, sd = 43860.05),
  Influenza.Deaths = rnorm(n, mean = 150.4058, sd = 826.2635)
)

# Ensure non-negative values for death counts
synthetic_data <- synthetic_data %>%
  mutate(
    Covid.19.Deaths = ifelse(Covid.19.Deaths < 0, 0, Covid.19.Deaths),
    Total.Deaths = ifelse(Total.Deaths < 0, 0, Total.Deaths),
    Pneumonia.Deaths = ifelse(Pneumonia.Deaths < 0, 0, Pneumonia.Deaths),
    Influenza.Deaths = ifelse(Influenza.Deaths < 0, 0, Influenza.Deaths)
  )

# Print the first few rows of the synthetic dataset
head(synthetic_data)
```


```{r}
#Histogram of Total Deaths
ggplot(synthetic_data, aes(x = Total.Deaths)) +
  geom_histogram(binwidth = 5000, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Total Deaths", x = "Total Deaths", y = "Frequency") +
  theme_minimal()
```

```{r}
#Frequency table for Age Group
age_group_table <- table(synthetic_data$Age.Group)
print(age_group_table)
```

```{r}
#Bar Plot of Age Groups
ggplot(synthetic_data, aes(x = Age.Group)) +
  geom_bar(fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Bar Plot of Age Groups", x = "Age Group", y = "Frequency") +
  theme_minimal()
```

```{r}
# Calculate percentage of counts by Sex
sex_counts <- synthetic_data %>%
  group_by(Sex) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)
```

```{r}
# Create the bar chart
ggplot(sex_counts, aes(x = Sex, y = Percentage, fill = Sex)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.7) +
  labs(title = "Percentage of Counts by Sex", x = "Sex", y = "Percentage (%)") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  theme_minimal() +
  theme(legend.position = "none")
```

